%!TEX root = ../main.tex
\section{Stack Machine}
\label{sec:stack-machine}

% - Hvad er en stack machine?
% - Udfordringer
% - Løsninger
% - Program representation
% - Compilation
% - Execution

Now that we can produce terms in our well-typed representation, we would like to go a step further. This means compiling these terms to some portable representation which can then be executed, while using dependent types to retain as much safety as possible. To accomplish this, we need a compiler to translate terms in the well-typed representation to instructions, and a machine to run these instructions. Due to its simplicity, we have chosen to implement a stack machine.

A minimal stack machine\,\cite[pp. 157--161]{Sestoft:PLC} needs only a few components: a list of instructions, a program counter (PC), a stack, and a stack pointer (SP). The program counter records how far into the list of instructions the program is, and the stack pointer points to the topmost element of the stack. As each instruction is executed, the stack is manipulated and the program counter is increased. Because each instruction operates on the top of the stack, programs are written in a postfix style. For example, to add the constants 3 and 4 together, the instructions required would be: \texttt{[PUSH 3, PUSH 4, ADD]}. The first two instructions push the constants onto the stack, and the third instruction adds them and leaves the result on the stack. Thus each \texttt{PUSH} instruction has the stack effect of adding one value to the stack, and the effect of the \texttt{ADD} instruction is to consume two values from the stack, and leave one value. The stack effect of the two instructions can be written (0, 1) and (2, 1), respectively.

\subsubsection{Problems}
In most stack machine implementations, many errors can occur at run time. For example, running the program \texttt{[PUSH 3, ADD]} with an empty stack will result in a stack underflow error, as the \texttt{ADD} instruction is expecting 2 values on the stack. It is up to the compiler to make sure all the instructions are correct. Unfortunately, most compilers are not accompanied by a proof of correctness. Instead they are thoroughly tested. As a consequence, even if you have proved the correctness of your program, as soon as you compile it with an unverified compiler, you lose all assurances, as Xavier Leroy points out in an article on the verified C compiler CompCert\,\cite{Leroy_formalverification}.

\subsection{Desired Qualities}
While it is beyond the scope of this project to produce a fully verified compiler and stack machine, we can use dependent types to make certain guarantees about the compiler, the programs it produces, and the stack machine that runs the programs. Specifically, we would like a program representation that encodes the stack effect of the instructions that comprise it, and a stack machine that only accepts \textit{well-stacked} programs that will run with an empty stack, without any stack underflow errors. The stack machine will be very simple. Instead of using a program counter and a stack pointer, the program and the stack themselves will be passed as parameters to the stack machine. As the stack machine is only used to run programs compiled from the well-typed representation, which has no terms requiring a heap, all values will be placed on the stack. This means a heap is not needed, and was not implemented.

\input{sections/program_representation}
\input{sections/compile}
\input{sections/running_a_program}
\input{sections/omissions}
